Beyond the Iceberg Hypothesis: \\Opening the Black Box of Transport Costs

### 2016-09-11 redaction/slides_ETSG1.tex
version définitive

\begin{frame}
\frametitle{Motivation (cont')}
\begin{itemize}
\item Standard modeling of trade costs: As an ad-valorem tax-equivalent \vspace{0.1cm}
\begin{itemize}
\item[-] As a constant percentage of the producer price per unit traded \vspace{0.1cm}
\item[$\Leftrightarrow$] The ``iceberg cost'' hypothesis (Samuelson, 1954) \vspace{0.1cm}
\end{itemize}
\item Yet... A debated question \vspace{0.1cm}
\item Would not trade costs rather exhibit an additive structure ?  \vspace{0.1cm}
\item The structure (additive vs iceberg) of transport costs, not anecdotal \vspace{0.1cm}
\begin{itemize}
\footnotesize{
\item[-] Additive trade costs \& the pattern of trade flows (Alchian \& Allen, 1964) \vspace{0.1cm}
\item[-] Strong normative implications (Sorensen, 2014)\vspace{0.1cm}
\item[-] The additive structure of trade costs is supported by recent empirical evidence (Hummels \& Skiba, 2004)  \vspace{0.1cm}}
\end{itemize}
\item[$\Rightarrow$] Trade costs are likely to display an additive component, but precisely... by how much? \vspace{0.1cm}
\end{itemize}
\textbf{One objective of the paper: Provide an answer to this question}
\end{frame}

\begin{frame}
\frametitle{Our paper in 3 questions (and 3 answers)}
An empirical decomposition of the structure of transport costs \vspace{0.1cm}
\begin{itemize}
\item[(1)] What is the size of the iceberg and the additive trade costs? \vspace{0.1cm}
\item[$\Rightarrow$] Provide a quantitative measure of both, using US imports data \vspace{0.1cm}
\begin{itemize}
\footnotesize{
\item[-] Iceberg cost: 2.5\% and 3.2\% of the export price in air and ocean transport (mean value over 1974-2013, US imports) \vspace{0.1cm}
\item[-] Additive cost: 1.8\% and 2.9\% of the export price  \vspace{0.1cm}
}
\end{itemize}
\item[(2)] What do we lose by skipping the additive part of transport costs? \vspace{0.1cm}
\item[$\Rightarrow$] We lose much: With the additive term included, \vspace{0.1cm}
\begin{itemize}
\item[-] The estimated iceberg component is reduced by a factor of 2 \vspace{0.1cm}
\item[-] A significantly better ``goodness-of-fit'' \vspace{0.1cm}
\end{itemize}
\item[(3)] How have international transport costs evolved over time? \vspace{0.1cm}
\begin{itemize}
\item[-] Transport costs \textit{per se} have fallen since 1985, by $\simeq$ 40\% \vspace{0.1cm}
\item[-] When additive costs are included, not much difference between air and sea transport, $\neq$ Hummels (2007) and Behar \& Venables (2010) \vspace{0.1cm}
%\item[-] Marked difference in the evolution of additive and multiplicative costs in Air
\end{itemize}
\end{itemize}
\end{frame}

@@

\begin{frame}
\frametitle{Empirical specification: The estimated equation}
\begin{itemize}
\item Relate the import price $p$ to the export price $\widetilde{p}$ given both additive (per-kg) costs $t$ and ad-valorem costs $\tau$:
$$p = \tau \widetilde{p} + t, \qquad \text{with}\quad \tau \geq 1,\quad t \geq 0$$
\item For product $k$, from country $i$  \vspace{0.1cm}
\item Rewrite to get:
$$\frac{p_{ik}}{\widetilde{p}_{ik}} -1 = \tau_{ik} -1 +\frac{t_{ik}}{ \widetilde{p}_{ik}}$$
\item[$\Rightarrow$] Estimate this equation for each year over 1974-2013 \vspace{0.1cm}
\begin{itemize}
\item[-] The equation being also year- and mode (air or vessel)- specific
\end{itemize}

@@

\begin{frame}[label=slide_results_summary]
\frametitle{Result 1: Estimating transport costs over time}
For both the ad-valorem and the additive components \vspace{0.1cm}
\begin{itemize}
\item[-] \footnotesize{Average values over 1974-2013, in percent of the export price \hyperlink{app_results_summary}{\beamergotobutton{More}}  } \vspace{0.1cm}
\end{itemize}
\begin{table}[htbp]
  \centering
  \scriptsize{
  %\caption{Transport costs estimates: Summary \label{tab:summary_results}}
  \begin{center}
    \begin{tabular}{l|cc|cc}
      \hline \hline
    %\multicolumn{5}{c}{Mean value over 1974-2013}   \\
    \# digit & \multicolumn{2}{c}{3 digits} & \multicolumn{2}{c}{4 digits} \\ \hline
    Mode  & Vessel & Air & Vessel & Air \\ \hline
    \multicolumn{5}{l}{With only Ad-Valorem Trade Costs ($\widehat{\tau}^{ice}$)}  \\ \hline
    Mean  & \bf{5.8} & \bf{5.1} & 6.0 & 4.9 \\
    Median & 5.1 & 4.2 & 5.2 & 3.7 \\ \hline
    %Std   & 0.032 & 0.042 & 0.036 & 0.045 \\
    %Min. value & 1.003 & 1.001 & 1.003 & 1.000 \\
    %Max. value & 1.304 & 1.685 & 1.408 & 2.051 \\ \hline
    \multicolumn{5}{l}{With Additive \& Ad-Valorem Trade Costs } \\ \hline
   \multicolumn{5}{l}{\textit{Ad-valorem term} ($\widehat{\tau}^{adv}$)}\\ \hline
    Mean  & \bf{3.2} & \bf{2.5} & 3.3 & 2.4 \\
    Median & 2.8 & 1.8 & 2.8 & 1.6 \\ \hline
    %Std   & 0.023 & 0.023 & 0.025 & 0.026 \\
    %Min. value & 1.001 & 1.000 & 1.000 & 1.000 \\
    %Max. value & 1.227 & 1.474 & 1.264 & 1.537 \\ \hline
    \multicolumn{5}{l}{\textit{Additive term }($\widehat{t}^{add}/\widetilde{p}$)} \\ \hline
    Mean  & \bf{2.9} & \bf{1.8} & 2.8 & 1.9 \\
    Median & 1.9 & 0.7 & 1.7 & 0.8 \\ \hline
    %Std   & 0.041 & 0.034 & 0.039 & 0.034 \\
    %Min. value & 0.000 & 0.000 & 0.000 & 0.000 \\
    %Max. value & 2.941 & 13.303 & 3.197 & 11.440 \\ \hline
        \# obs. & 29279 & 28207 & 29317 & 27680 \\ \hline \hline
   % \# origin country & 188 & 191 & 188 & 189 \\
   % \# products & 230 & 211 & 666 & 567 \\
  \end{tabular}
%\parbox[l]{8cm}{\tiny{Notes: Statistics are obtained weighting each observation by its value.}}
%. The additive term is expressed in fraction of fab price. ($^\ast$): Four 4-digit estimation: 0n selected years. ($^{\ast \ast}$): 1989 omitted in 3 digit estimation for air.}}
\end{center}}
  \end{table}%
\begin{itemize}
\item Transport costs are sizeable \vspace{0.1cm}
\begin{itemize}
\item[-] Add $\simeq$ a 5\% margin over the export price \vspace{0.1cm}
\end{itemize}
\item Opening the black box of transport costs \vspace{0.1cm}
\begin{itemize}
\item[-] Iceberg cost: 2.5\% and 3.2\% of the export price in Air \& Vessel resp.
\item[-] Additive cost: 1.8\% and 2.9\% of the export price
\end{itemize}
\end{itemize}
\end{frame}

@@

\begin{frame}[label=app_method_1]
\frametitle{More on the estimation strategy (1)}
\begin{itemize}
\item Assumptions on the specification of transport costs (as in Irrarazabal et al., 2015) \vspace{0.1cm}
\begin{itemize}
\item[-] Both iceberg and additive costs are separable between the origin country $i$ and the product $k$ dimensions \vspace{0.1cm}
\item[-] Separability in a multiplicative manner for ad-valorem costs and additive manner for per-kg costs \vspace{0.1cm}
\item[$\Leftrightarrow$] Write $t_{ik}$ and $\tau_{ik}$ as:
\begin{equation}
 \tau_{ik} = \tau_{i} \times \tau_{k}, \qquad t_{ik} = t_{i} + t_{k} \label{eq:specifTC}
 \end{equation}
\end{itemize}
\item Given the constraint $\frac{p_{ik}}{\widetilde{p}_{ik}} -1$, the error term should be always positive and multiplicative \vspace{0.1cm}
\item[$\Rightarrow$] The estimated equation becomes:
$$\frac{p_{ik}}{\widetilde{p}_{ik}}-1 =\left(\tau_{i} \times \tau_{k} -1+\frac{t_{i} + t_{k}}{\widetilde{p}_{ik}} \right)\times \exp(\epsilon_{ik})$$
\begin{itemize}
\item[-] With $\epsilon_{ik}$ following a normal law centered on 0.
\end{itemize}
\end{itemize}
\hyperlink{slide_method}{\beamergotobutton{Back to slide}}
\end{frame}

\begin{frame}[label=app_method_2]
\frametitle{More on the estimation strategy (2)}
\begin{itemize}
\item The non-linear least squares method \vspace{0.1cm}
\begin{itemize}
\item[-] At the basis of the method: Approximate the model by a linear one and refine the parameters by successive iterations \vspace{0.1cm}
\item[-] The criterion for convergence: That the sum of the squares of the residuals does not not increase? from one iteration to the next \vspace{0.1cm}
\end{itemize}
\item Eliminate potential influence of outliers: Exclude the 5 percent of the upper and lower tails of the distribution \vspace{0.1cm}
\item Obtaining the final estimates: More details  \vspace{0.1cm}
\begin{itemize}
\item[-] After estimating Equations (\ref{eq:model_with_add}) and (\ref{eq:model_nlI}), re-built \vspace{0.1cm}
\begin{itemize}
\item[$\star$] With additive costs:
$$\widehat{\tau}^{adv}_{ik} = \widehat{\tau_{i}} \times \widehat{\tau_{k}}, \qquad \widehat{t}^{add}_{ik} = \widehat{t}_{i} + \widehat{t}_{k}$$
\item[$\star$] With only iceberg costs:
$$\widehat{\tau}^{ice}_{ik} = \widehat{\tau_{i}} \times \widehat{\tau_{k}}$$
\end{itemize}
\item[-] Then, take the weighted average by country-product, \vspace{0.1cm}
\item[-] Using the values of each trade flow ($ik$-specific) over total yearly trade as a weighting scheme
\end{itemize}
\end{itemize}
\hyperlink{slide_method}{\beamergotobutton{Back to slide}}
\end{frame}

>>>>

\begin{frame}
\frametitle{Motivation (cont')}
\begin{itemize}
\item Standard modeling of trade costs: As an ad-valorem tax-equivalent \vspace{0.1cm}
\begin{itemize}
\item[-] As a constant percentage of the producer price per unit traded \vspace{0.1cm}
\item[$\Leftrightarrow$] The ``iceberg cost'' hypothesis (Samuelson, 1954) \vspace{0.1cm}
\end{itemize}
\item Yet... A debated question \vspace{0.1cm}
\item Would not trade costs rather exhibit an additive structure ?  \vspace{0.1cm}
\item The structure (additive vs iceberg) of transport costs, not anecdotal \vspace{0.1cm}
\begin{itemize}
\footnotesize{
\item[-] Additive trade costs \& the pattern of trade flows (Alchian \& Allen, 1964) \vspace{0.1cm}
\item[-] Strong normative implications (Sorensen, 2014)\vspace{0.1cm}
\item[-] The additive structure of trade costs is supported by recent empirical evidence (Irrarazabal et al. 2014, Hummels \& Skiba, 2004)  \vspace{0.1cm}}
\end{itemize}
\item[$\Rightarrow$] Trade costs are likely to display an additive component, but precisely... by how much? \vspace{0.1cm}
\end{itemize}
\textbf{One objective of the paper: Provide an answer to this question}
\end{frame}

\begin{frame}
\frametitle{Our paper in 3 questions (and 3 answers)}
An empirical decomposition of the structure of transport costs \vspace{0.1cm}
\begin{itemize}
\item[(1)] What is the size of the iceberg and the additive transport costs? \vspace{0.1cm}
\item[$\Rightarrow$] Provide a quantitative measure of both, using US imports data \vspace{0.1cm}
\begin{itemize}
\footnotesize{
\item[-] Iceberg cost: 2.5\% and 3.2\% of the export price in air and ocean transport (mean value over 1974-2013, US imports) \vspace{0.1cm}
\item[-] Additive cost: 1.8\% and 2.9\% of the export price  \vspace{0.1cm}
}
\end{itemize}
\item[(2)] What do we lose by skipping the additive part of transport costs? \vspace{0.1cm}
\item[$\Rightarrow$] We lose much: With the additive term included, \vspace{0.1cm}
\begin{itemize}
\item[-] The estimated iceberg component is reduced by a factor of 2 \vspace{0.1cm}
\item[-] A significantly better ``goodness-of-fit'' \vspace{0.1cm}
\end{itemize}
\item[(3)] How have international transport costs evolved over time? \vspace{0.1cm}
\begin{itemize}
\item[-] Transport costs \textit{per se} have fallen since 1985, by $\simeq$ 40\% \vspace{0.1cm}
\item[-] When additive costs are included, not much difference between air and sea transport, $\neq$ Hummels (2007) and Behar \& Venables (2010) \vspace{0.1cm}
%\item[-] Marked difference in the evolution of additive and multiplicative costs in Air
\end{itemize}
\end{itemize}
\end{frame}

@@

\begin{frame}
\frametitle{Empirical specification: The estimated equation}
\begin{itemize}
\item Relate the import price $p$ to the export price $\widetilde{p}$ given both additive (per-kg) costs $t$ and ad-valorem costs $\tau$:
$$p = \tau \widetilde{p} + t, \quad \text{with }~~\tau \geq 1, \quad t \geq 0$$
\item For product $k$, from country $i$  \vspace{0.1cm}
\item Rewrite to get:
$$\frac{p_{ik}}{\widetilde{p}_{ik}} -1 = \tau_{ik} -1 +\frac{t_{ik}}{ \widetilde{p}_{ik}}$$
\item[$\Rightarrow$] Estimate this equation for each year over 1974-2013 \vspace{0.1cm}
\begin{itemize}
\item[-] The equation being also year- and mode (air or vessel)- specific
\end{itemize}

@@

\begin{frame}[label=slide_results_summary]
\frametitle{Result 1: Estimating transport costs over time}
For both the ad-valorem and the additive components \vspace{0.1cm}
\begin{itemize}
\item[-] \footnotesize{Average values over 1974-2013, in percent of the export price \hyperlink{app_results_summary}{\beamergotobutton{More}}  } \vspace{0.1cm}
\end{itemize}
\begin{table}[htbp]
  \centering
  \scriptsize{
  %\caption{Transport costs estimates: Summary \label{tab:summary_results}}
  \begin{center}
    \begin{tabular}{l|cc|cc}
      \hline \hline
    %\multicolumn{5}{c}{Mean value over 1974-2013}   \\
    \# digit & \multicolumn{2}{c}{3 digits} & \multicolumn{2}{c}{4 digits} \\ \hline
    Mode  & Vessel & Air & Vessel & Air \\ \hline
    \multicolumn{5}{l}{With only Ad-Valorem Costs ($\widehat{\tau}^{ice}$)}  \\ \hline
    Mean  & \bf{5.8} & \bf{5.1} & 6.0 & 4.9 \\
    Median & 5.1 & 4.2 & 5.2 & 3.7 \\ \hline
    %Std   & 0.032 & 0.042 & 0.036 & 0.045 \\
    %Min. value & 1.003 & 1.001 & 1.003 & 1.000 \\
    %Max. value & 1.304 & 1.685 & 1.408 & 2.051 \\ \hline
    \multicolumn{5}{l}{With Additive \& Ad-Valorem Costs } \\ \hline
   \multicolumn{5}{l}{\textit{Ad-valorem term} ($\widehat{\tau}^{adv}$)}\\ \hline
    Mean  & \bf{3.2} & \bf{2.5} & 3.3 & 2.4 \\
    Median & 2.8 & 1.8 & 2.8 & 1.6 \\ \hline
    %Std   & 0.023 & 0.023 & 0.025 & 0.026 \\
    %Min. value & 1.001 & 1.000 & 1.000 & 1.000 \\
    %Max. value & 1.227 & 1.474 & 1.264 & 1.537 \\ \hline
    \multicolumn{5}{l}{\textit{Additive term }($\widehat{t}^{add}/\widetilde{p}$)} \\ \hline
    Mean  & \bf{2.9} & \bf{1.8} & 2.8 & 1.9 \\
    Median & 1.9 & 0.7 & 1.7 & 0.8 \\ \hline
    %Std   & 0.041 & 0.034 & 0.039 & 0.034 \\
    %Min. value & 0.000 & 0.000 & 0.000 & 0.000 \\
    %Max. value & 2.941 & 13.303 & 3.197 & 11.440 \\ \hline
        \# obs. & 29279 & 28207 & 29317 & 27680 \\ \hline \hline
   % \# origin country & 188 & 191 & 188 & 189 \\
   % \# products & 230 & 211 & 666 & 567 \\
  \end{tabular}
%\parbox[l]{8cm}{\tiny{Notes: Statistics are obtained weighting each observation by its value.}}
%. The additive term is expressed in fraction of fab price. ($^\ast$): Four 4-digit estimation: 0n selected years. ($^{\ast \ast}$): 1989 omitted in 3 digit estimation for air.}}
\end{center}}
  \end{table}%
\begin{itemize}
\item Transport costs are sizeable \vspace{0.1cm}
\begin{itemize}
\item[-] Add $\simeq$ a 5\% margin over the export price \vspace{0.1cm}
\end{itemize}
\item Opening the black box of transport costs \vspace{0.1cm}
\begin{itemize}
\item[-] Iceberg cost: 2.5\% and 3.2\% of the export price in Air \& Vessel resp.
\item[-] Additive cost: 1.8\% and 2.9\% of the export price
\end{itemize}
\end{itemize}
\end{frame}

@@

\begin{frame}[label=app_method_1]
\frametitle{More on the estimation strategy (1)}
\begin{itemize}
\item Assumptions on the specification of transport costs (as in Irrarazabal et al., 2015) \vspace{0.1cm}
\begin{itemize}
\item[-] Both iceberg and additive costs are separable between the origin country $i$ and the product $k$ dimensions \vspace{0.1cm}
\item[-] Separability in a multiplicative manner for ad-valorem costs and additive manner for per-kg costs \vspace{0.1cm}
\item[$\Leftrightarrow$] Write $t_{ik}$ and $\tau_{ik}$ as:
\begin{equation}
 \tau_{ik} = \tau_{i} \times \tau_{k}, \qquad t_{ik} = t_{i} + t_{k} \label{eq:specifTC}
 \end{equation}
\end{itemize}
\item Given the constraint $\frac{p_{ik}}{\widetilde{p}_{ik}} \geq 1$, the error term should be always positive and multiplicative \vspace{0.1cm}
\item[$\Rightarrow$] The estimated equation becomes:
$$\frac{p_{ik}}{\widetilde{p}_{ik}}-1 =\left(\tau_{i} \times \tau_{k} -1+\frac{t_{i} + t_{k}}{\widetilde{p}_{ik}} \right)\times \exp(\epsilon_{ik})$$
\begin{itemize}
\item[-] With $\epsilon_{ik}$ following a normal law centered on 0.
\end{itemize}
\end{itemize}
\hyperlink{slide_method}{\beamergotobutton{Back to slide}}
\end{frame}

\begin{frame}[label=app_method_2]
\frametitle{More on the estimation strategy (2)}
\begin{itemize}
\item The non-linear least squares method \vspace{0.1cm}
\begin{itemize}
\item[-] At the basis of the method: Approximate the model by a linear one and refine the parameters by successive iterations \vspace{0.1cm}
\item[-] The criterion for convergence: That the sum of the squares of the residuals does not increase from one iteration to the next \vspace{0.1cm}
\end{itemize}
\item Eliminate potential influence of outliers: Exclude the 5 \% of the upper and lower tails of the distribution \vspace{0.1cm}
\item Obtaining the final estimates: More details  \vspace{0.1cm}
\begin{itemize}
\item[-] After estimating Equations (\ref{eq:model_with_add}) and (\ref{eq:model_nlI}), re-built \vspace{0.1cm}
\begin{itemize}
\item[$\star$] With additive costs:
$$\widehat{\tau}^{adv}_{ik} = \widehat{\tau_{i}} \times \widehat{\tau_{k}}, \qquad \widehat{t}^{add}_{ik} = \widehat{t}_{i} + \widehat{t}_{k}$$
\item[$\star$] With only iceberg costs:
$$\widehat{\tau}^{ice}_{ik} = \widehat{\tau_{i}} \times \widehat{\tau_{k}}$$
\end{itemize}
\item[-] Then, take the weighted average by country-product, \vspace{0.1cm}
\item[-] Using the values of each trade flow ($ik$-specific) over total yearly trade as a weighting scheme
\end{itemize}
\end{itemize}
\hyperlink{slide_method}{\beamergotobutton{Back to slide}}
\end{frame}
